Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-06-05 paper/paper.qmd
Include title & authors in markdown

```{r abstract}
#| output: asis
if ('abstract' %in% names(rmarkdown::metadata)) {
  cat('# Abstract\n', rmarkdown::metadata$abstract)
}
if ('importance' %in% names(rmarkdown::metadata)) {
  cat('# Importance\n', rmarkdown::metadata$importance)
}
```
:::

@@

There is a need for robust, accurate methods to identify patients at risk of
severe CDI outcomes.
When paired with treatment options that may reduce risk of severity, prediction
models can guide clinician decision-making to improve patient outcomes while
minimizing harms from unnecessary treatment.
Numerous scoring systems for predicting severe CDI outcomes based on patient
clinical factors have been developed, but none have validated well to external
datasets nor are any in use in routine clinical practice
[@chakra_prediction_2012;@perry_external_2022].
Machine learning (ML) is a promising approach that allows for use of thousands
of features to  classify samples and predict outcomes, rather than relying on
limited sets of human-curated variables.
Indeed, ML models trained on entire Electronic Health Record (EHR) data
have demonstrated improved performance over curated models
[@rao_clostridium_2015;@li_using_2019;@dieterle_systemic_2020].

@@

There is not currently a consensus definition of CDI severity.
Some scoring systems leverage clinical data available during the course of CDI,
while others focus on adverse outcomes of CDI at 30 days after diagnosis
[@ressler_defining_2021].
We explored four different ways to define CDI cases as severe or not (@fig-flowchart).
The "IDSA" definition of severe CDI is based on laboratory values collected on
the day of diagnosis, with a case being severe if serum creatinine level is
greater than or equal to $1.5 mg/dL$ and the white blood cell count is greater
than or equal to $15 k/\mu L$ [@mcdonald_clinical_2018].
However, the IDSA score is known to be a poor predictor of adverse outcomes
[@stevens_validation_2020], although it is straightforward to collect.
The remaining definitions we employed focus on the occurrence of adverse
outcomes, which may be more clinically relevant.
The "attributable" severity definition is based on disease-related complications
defined by the CDC, where an adverse event of ICU admission, colectomy, or death
occurs within 30 days of CDI diagnosis, and the adverse event is determined to
be attributable to the CDI by physician chart review [@mcdonald_recommendations_2007].
However, physician chart review is time-consuming and has not been completed for
all cases, so we defined "all-cause" severity where a case is severe if an
adverse event occurs within 30 days of the diagnosis regardless of the cause of
the adverse event.
Finally, we defined a "pragmatic" severity definition that makes use of the
attributable definition when available and falls back to the all-cause
definition when chart review has not been completed, allowing us to use as many
samples as we have available while taking physicians' expert opinions into
account where possible (@fig-flowchart B).
We trained ML models to classify (in the case of the IDSA definition) or predict
(in the case of the three other definitions) severity and determined how well
OTU-based models perform for each definition.

>>>>

```{r abstract}
#| output: asis
cat('# ', rmarkdown::metadata$title, '\n')
cat(paste(sapply(rmarkdown::metadata$author, 
                 function(x) {x$name}), 
          collapse = ', '
          ), 
    '\n\n'
)
if ('abstract' %in% names(rmarkdown::metadata)) {
  cat('# Abstract\n', rmarkdown::metadata$abstract)
}
if ('importance' %in% names(rmarkdown::metadata)) {
  cat('# Importance\n', rmarkdown::metadata$importance)
}
```
:::

@@

There is a need for robust, accurate methods to identify patients at risk of
severe CDI outcomes.
When paired with treatment options that may reduce risk of severity, prediction
models can guide clinician decision-making to improve patient outcomes while
minimizing harms from unnecessary treatment.
Numerous scoring systems for predicting severe CDI outcomes based on patient
clinical factors have been developed, but none have validated well to external
datasets nor are any in use in routine clinical practice
[@chakra_prediction_2012;@perry_external_2022].
Machine learning (ML) is a promising approach that allows for use of thousands
of features to classify samples and predict outcomes, rather than relying on
limited sets of human-curated variables.
Indeed, ML models trained on entire Electronic Health Record (EHR) data
have demonstrated improved performance over curated models
[@rao_clostridium_2015;@li_using_2019].

@@

There is not currently a consensus definition of CDI severity.
Some scoring systems leverage clinical data available during the course of CDI,
while others focus on adverse outcomes of CDI at 30 days after diagnosis
[@ressler_defining_2021;@dieterle_systemic_2020].
We explored four different ways to define CDI cases as severe or not (@fig-flowchart).
The "IDSA" definition of severe CDI is based on laboratory values collected on
the day of diagnosis, with a case being severe if serum creatinine level is
greater than or equal to $1.5 mg/dL$ and the white blood cell count is greater
than or equal to $15 k/\mu L$ [@mcdonald_clinical_2018].
However, the IDSA score is known to be a poor predictor of adverse outcomes
[@stevens_validation_2020], although it is straightforward to collect.
The remaining definitions we employed focus on the occurrence of adverse
outcomes, which may be more clinically relevant.
The "attributable" severity definition is based on disease-related complications
defined by the CDC, where an adverse event of ICU admission, colectomy, or death
occurs within 30 days of CDI diagnosis, and the adverse event is determined to
be attributable to the CDI by physician chart review [@mcdonald_recommendations_2007].
However, physician chart review is time-consuming and has not been completed for
all cases, so we defined "all-cause" severity where a case is severe if an
adverse event occurs within 30 days of the diagnosis regardless of the cause of
the adverse event.
Finally, we defined a "pragmatic" severity definition that makes use of the
attributable definition when available and falls back to the all-cause
definition when chart review has not been completed, allowing us to use as many
samples as we have available while taking physicians' expert opinions into
account where possible (@fig-flowchart B).
We trained ML models to classify (in the case of the IDSA definition) or predict
(in the case of the three other definitions) severity and determined how well
OTU-based models perform for each definition.
